---
title: "Getting started with datagrade"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with datagrade}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 4.5, dpi = 120)
library(datagrade)
```

## One call

`dg_assess()` takes a data frame or a file path and returns a `dg_report`
object. Nothing is written to the global environment, and the data frame you
pass in is never modified.

```{r}
report <- dg_assess(iris, verbose = FALSE)
report
```

## The object, not the printout

The reference implementation this package grew out of printed everything with
`cat()` and returned `NULL`, so nothing could be used programmatically. Here the
printout is a side effect and the object is the product.

```{r}
report$scores
report$outliers$by_column
report$redundancy$pairs
```

`summary()` gives the same content as tables:

```{r}
summary(report)$columns
```

## Assessing many data sets at once

`as.data.frame()` collapses a report to one row, which is what makes a corpus
sweep a two-liner:

```{r}
datasets <- list(iris = iris, mtcars = mtcars, airquality = airquality)
sweep <- do.call(rbind, lapply(names(datasets), function(nm) {
  row <- as.data.frame(dg_assess(datasets[[nm]], verbose = FALSE))
  row$source <- nm
  row
}))
sweep[, c("source", "rows", "completeness", "accuracy", "consistency", "overall")]
```

## Figures

`dg_plots()` returns a named list of ggplot objects. Every figure is built from
statistics computed during the assessment, so the plot objects never carry the
raw observations.

```{r}
figures <- dg_plots(report)
names(figures)
figures$scores
```

```{r}
figures$correlation
```

```{r, fig.height = 5.5}
figures$distribution
```

Use `dg_save_plots(report, "figures/")` to write them all to disk.

## Reading from a file

```{r}
path <- tempfile(fileext = ".csv")
write.csv(mtcars, path, row.names = FALSE)
dg_assess(path = path, verbose = FALSE)$scores
unlink(path)
```

`.csv`, `.tsv` and `.txt` are read with base R. `.xls` and `.xlsx` need the
suggested **readxl** package.

## Helpers stand alone

Each of the checks is exported and usable on its own:

```{r}
detect_data_types(data.frame(id = c("a", "b", "c"), n = c(1.5, 2.5, 3.5)))
dg_vif(iris)
analyze_distribution(iris)
```

## Where to next

* `vignette("methodology")` -- every formula, and how each score is computed.
* `vignette("scalability")` -- measured behaviour from 54 thousand to 10 million rows.
* `vignette("deviations")` -- how this package differs from the dissertation it implements, and why.
