---
title: "Methodology: every formula, stated"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodology: every formula, stated}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(datagrade)
```

Every score this package reports is written out below. Nothing is a black box,
and where a choice is arguable the argument is given rather than hidden.

All four dimension scores are **ratios in $[0, 1]$**, the form ISO/IEC 25024
measures take. `as_pct()` converts once, at the point of display, to 0--100.
Holding one scale internally is deliberate: the implementation this package
replaces kept completeness on 0--10 and consistency on 0--1 and multiplied both
by ten, which is why its published consistency figure for `iris` reads 2.5%
where 25% was meant.

## Completeness

$$
\text{completeness} = \left(1 - \frac{1}{p}\sum_{j=1}^{p} \frac{m_j}{n}\right)
$$

where $m_j$ is the number of missing values in column $j$, $n$ the row count and
$p$ the column count.

This is the **mean of the per-column missing rates**, not the overall share of
missing cells. The two differ whenever columns are unequally incomplete, and the
column-averaged form dilutes a badly damaged column: one column that is 30%
empty out of four costs 0.075, not 0.3. That is a property of the published
formula, kept here for continuity with the dissertation; if you want the cell
ratio, compute it from `report$missing$columns`.

```{r}
report <- dg_assess(airquality, verbose = FALSE)
report$missing$columns
report$scores[["completeness"]]
```

## Accuracy

$$
\text{accuracy} = \left(1 - \frac{\text{outliers flagged}}{\text{numeric values tested}}\right)
$$

Outliers are flagged by one of two rules.

**Z-score.** A value is flagged when $\left|\frac{x_i - \mu}{\sigma}\right| > 3$.

**IQR.** A value is flagged when it falls outside
$[Q_1 - 1.5\,\text{IQR},\; Q_3 + 1.5\,\text{IQR}]$.

Two implementation choices are worth stating plainly.

*Outliers are counted on observed values, not imputed ones.* Mean imputation
adds mass at the centre of the distribution and shrinks $\sigma$, which
mechanically suppresses the very z-scores this step exists to find. The
dissertation ran this step after imputation; this package does not.

*A zero-variance column contributes zero outliers, not `NA`.* `scale()` returns
`NaN` when $\sigma = 0$, and summing that propagated `NA` through the entire
report, so a single constant column made the whole outlier count unreportable.

```{r}
df <- data.frame(constant = rep(5, 50), varying = c(rnorm(49), 100))
detect_outliers(df)$by_column
```

## Consistency

$$
\text{consistency} = \frac{\left|\{j : \mathrm{VIF}_j \le \tau\}\right|}{p_{\text{num}}}
$$

with $\tau$ = `vif_threshold`, default 5.

The variance inflation factor is defined as $\mathrm{VIF}_j = \frac{1}{1 - R_j^2}$,
where $R_j^2$ comes from regressing column $j$ on all other numeric columns. This
package does not fit those regressions. It uses the identity that the VIFs are
the **diagonal of the inverted correlation matrix**:

$$
\mathrm{VIF} = \operatorname{diag}\left(\mathbf{R}^{-1}\right)
$$

which gives the same numbers in $O(p^3)$ time, independent of the row count. The
equivalence is asserted in the test suite against explicit `lm()` fits:

```{r}
set.seed(11)
d <- data.frame(a = rnorm(500), b = rnorm(500))
d$c <- 0.6 * d$a + 0.3 * d$b + rnorm(500)

dg_vif(d)

vapply(names(d), function(col) {
  r2 <- summary(lm(d[[col]] ~ ., data = d[setdiff(names(d), col)]))$r.squared
  1 / (1 - r2)
}, numeric(1))
```

Perfectly collinear columns are reported as `Inf` rather than dropped, so they
stay visible and still count against the score.

## Timeliness

$$
\text{timeliness} = \left(1 - \frac{\text{future dates} + \text{stale dates}}{\text{dates present}}\right)
$$

A date is *future* if it is after `Sys.Date()`, *stale* if it is more than
`stale_years` (default 10) old.

This dimension is only measurable if date columns exist. Character columns are
therefore parsed as dates **before** any text normalisation runs -- otherwise
punctuation stripping turns `"2024-01-15"` into `"20240115"` and no date column
ever exists.

```{r}
df <- data.frame(when = c("2024-01-15", "2024-02-20"), value = 1:2)
r <- dg_assess(df, verbose = FALSE, strip_punctuation = TRUE)
r$parsed_dates
r$dates
```

## Distribution

Skewness and kurtosis use the sample moment forms:

$$
g_1 = \frac{1}{n}\sum \left(\frac{x_i-\mu}{\sigma}\right)^3
\qquad
g_2 = \frac{1}{n}\sum \left(\frac{x_i-\mu}{\sigma}\right)^4 - 3
$$

The $-3$ makes $g_2$ **excess** kurtosis, so a normal distribution scores 0. This
matters: the dissertation's formula subtracts 3 but its code called
`moments::kurtosis()`, which does not, so the published formula and the
published code disagreed by exactly 3. The formula is the contract here.

Normality uses the Shapiro-Wilk test, which is undefined above 5,000
observations. Larger columns are tested on a deterministic subsample and the
size actually used is reported in `shapiro_n`, so the result is never presented
as though the whole column were tested.

```{r}
set.seed(42)
analyze_distribution(data.frame(normal = rnorm(20000), skewed = rexp(20000)))
```

## The overall score

$$
\text{overall} = \frac{\sum_{d \in D} w_d \, s_d}{\sum_{d \in D} w_d}
$$

where $D$ is the set of dimensions that **apply** to the data. A data set with
no date columns cannot be scored on timeliness, so timeliness is dropped and the
remaining weights are renormalised rather than the data being penalised for a
dimension it cannot have.

```{r}
text_only <- data.frame(a = c("x", "y", "z"), stringsAsFactors = FALSE)
r <- dg_assess(text_only, verbose = FALSE)
r$scores
r$overall
```

## The ISO/IEC 25012 model

Everything above is a statistic. This section is how those statistics become
scores against the standard's model.

ISO/IEC 25012 defines five **inherent characteristics** of data quality --
accuracy, completeness, consistency, credibility and currentness -- and beneath
them fifteen **quality properties**. ISO/IEC 25024 gives each property a
measurement function of the form

$$
X = \frac{A}{B}
$$

on a ratio scale over $[0.0, 1.0]$, where $A$ and $B$ are counts of *quality
measure elements*: $A$ the items satisfying the property, $B$ the items **for
which the property is defined at all**.

That second count is the whole difficulty. For the accuracy-range property,
$B$ is "the number of data items for which can be defined a required interval of
values" -- and a required interval is not in the data. It is declared by whoever
knows what the data means. This is exactly why every other R package in this
space demands a data dictionary before it will run.

`datagrade` splits the difference. Six properties have a $B$ countable from the
table itself and are always measured; nine have a $B$ that only a declaration
can supply, and `dg_spec()` is where it goes.

```{r}
report <- dg_assess(iris, verbose = FALSE)
report$measures[, c("code", "characteristic", "value", "a", "b")]
```

### The six that need no declaration

**`COMP_REG`, record completeness.** $A$ is records with every required field
populated, $B$ is all records. With no `required` in the spec, every column
counts as required.

**`COMP_VAL_ESP`, data value completeness.** $A$ is populated cells in the
required columns, $B$ is all cells in them. This differs from the dissertation's
completeness, which averaged the per-column missing rates and so diluted a badly
damaged column; the cell ratio is the form the standard's $A/B$ takes.

**`FAL_COMP_FICH`, false completeness of file.** A cell can be occupied and
still empty. $A$ is text values that are not placeholders, $B$ is all non-null
text values in the required columns. The placeholder list is short and
conservative on purpose -- every entry on it (`"none"`, `"unknown"`, `"-999"`)
is a legitimate value in *some* data set, and a false positive silently lowers
the score. Matching is case-insensitive after trimming. Replace the list with
`null_equivalents`, or pass `character(0)` to disable the check.

**`RIES_INCO`, risk of inconsistency.** $A$ is numeric columns with
$\mathrm{VIF} \le \tau$, $B$ is numeric columns with a computable VIF. The
reasoning: information held more than once can contradict itself, and a column
whose variance is largely explained by the others carries no independent
content. This is a **proxy**, not a contradiction count -- the standard's
property concerns data items at risk of inconsistency, and collinearity is the
only signal for that available without declared rules. Declared rules go to
`CONS_SEMAN` instead.

**`CONS_FORM`, format consistency.** Each value is reduced to a *signature*:
runs of digits become `9`, runs of letters `A`, everything else is kept. So
`"2024-01-15"` and `"1999-12-31"` share the signature `9-9-9`. $A$ is values
matching their column's dominant signature, $B$ is non-null text values.

The trap is free text. Every sentence is its own signature, so a prose column
would score near zero for failing to conform to a format it never had. A column
is therefore only measured when its values collapse to a handful of shapes --
at most $\max(3, \lceil 0.05 \, d \rceil)$ signatures for $d$ distinct values.
Columns that fail this test are dropped from $B$ and named in the measure's
`note`, not scored.

```{r}
d <- data.frame(ref = c(rep("AB-1234", 8), "1234-AB", "XX/99"),
                stringsAsFactors = FALSE)
m <- dg_assess(d, verbose = FALSE)$measures
m[m$code == "CONS_FORM", c("value", "a", "b")]
```

Declaring `formats` in the spec replaces the inferred signature with a required
one, which changes the question from "is this column internally consistent?" to
"does it match the rule?".

**`CONV_ACT`, timeliness of update.** $A$ is dates neither in the future nor
older than `max_age` (ten years by default), $B$ is all dates present. Future
dates are counted against the score rather than ignored: they are impossible,
and treating them as fresh would let a corrupt timestamp *raise* currentness.

### The nine that need a declaration

Each is `NA` until the corresponding `dg_spec()` argument is given, and each
carries a `note` saying which argument that is. `NA` here is not a bad score --
it is the statement that $B$ could not be counted. Aggregation drops these
rather than scoring them zero, so an undocumented data set is never punished for
being undocumented.

Two details are worth stating because they change what the denominators mean.

*`EXAC_SEMAN` joins on the declared key first.* A record with no counterpart in
the reference is unverifiable, not inaccurate, so it never enters $B$.

*`CONS_SEMAN` counts rule evaluations, not records.* $B$ is records times
evaluable rules. A rule that errors -- because it names a column the data does
not have -- is a specification error rather than a data failure: it is excluded
from $B$ and named in the `note`.

```{r}
d <- data.frame(x = 1:4)
m <- dg_assess(d, spec = dg_spec(rules = list(fine = ~ x > 0,
                                              broken = ~ missing_col > 0)),
               verbose = FALSE)$measures
m[m$code == "CONS_SEMAN", c("value", "a", "b", "note")]
```

### Plausibility, which is not accuracy

The dissertation's accuracy score flagged outliers against an interval derived
from the data -- $\mu \pm 3\sigma$, or the IQR fence. All three ISO accuracy
properties require an interval, pattern or reference that was *declared*.
Substituting a derived one measures something real but different: an outlier
need not be wrong, and a wrong value can sit exactly on the mean.

So the measure is kept, renamed `plausibility`, reported beside the five
characteristics and excluded from `overall`. It is the most useful number in the
report for data that arrives with no documentation, and it is not the standard's
measure.

```{r}
report$plausibility
report$scores[["accuracy"]]   # NA: no interval was declared
```

### Aggregation, which the standard does not specify

Properties combine into a characteristic by the unweighted mean over those that
apply; characteristics combine into `overall` by the weighted mean over those
that apply. **ISO/IEC 25024 does not address how measures should be
aggregated**, so both steps are this package's choice and are stated as such
rather than presented as conformance.

$$
\text{characteristic} = \frac{1}{|P|}\sum_{p \in P} X_p
\qquad
\text{overall} = \frac{\sum_{c \in C} w_c \, s_c}{\sum_{c \in C} w_c}
$$

where $P$ is the applicable properties of that characteristic and $C$ the
applicable characteristics.

### What is still not claimed

The *model* implemented here -- five characteristics, fifteen properties, their
identifiers, the $A/B$ form -- is sourced below. What is **not** verified
against the standard is the exact definition of each property's quality measure
elements: precisely which items ISO counts into $A$ and $B$. Those live in
Annexes A--E of ISO/IEC 25024, which are not in any public source. Where this
package had to choose, the choice is written out above and visible in the `a`
and `b` columns of `report$measures`, so it can be checked and disagreed with.

Note also that ISO's word is **currentness**; *timeliness* comes from a
different tradition and was the dissertation's word. It is kept as
`report$timeliness` and accepted in `weights`.

### Sources for the claims in this section

ISO/IEC 25012 and 25024 are paywalled. Every statement above about their
*structure* is cited below. The standard's own measurement functions are
deliberately **not** reproduced, because they could not be verified without the
documents themselves.

* ISO/IEC 25024:2015. *Systems and software engineering — Systems and software
  Quality Requirements and Evaluation (SQuaRE) — Measurement of data quality.*
  <https://www.iso.org/standard/35749.html>
* ISO/IEC 25012:2008. *Software engineering — SQuaRE — Data quality model.*
  <https://www.iso.org/standard/35736.html>
* Gualo, F., Rodriguez, M., Verdugo, J., Caballero, I. and Piattini, M. (2021).
  'Data Quality Certification using ISO/IEC 25012: Industrial Experiences',
  *Journal of Systems and Software*. <https://arxiv.org/abs/2102.11527>

    The source for three specific claims: that ISO/IEC 25024 measures take the
    form $X = A/B$ on a ratio scale over $[0.0, 1.0]$; that the Accuracy Range
    measure (RAN_EXAC) defines its element $B$ as "the number of data items for
    which can be defined a required interval of values"; and that "ISO/IEC 25024
    does not specifically address how the measures corresponding to the data
    quality properties should be aggregated to compute the overall quality level
    for each data quality characteristic." The authors are from AQCLab, an
    accredited ISO/IEC 25012 data quality evaluation laboratory.

* Calabrese, J., Esponda, S. and Pesado, P. (2020). 'Framework for Data Quality
  Evaluation Based on ISO/IEC 25012 and ISO/IEC 25024', *8th Conference on Cloud
  Computing, Big Data & Emerging Topics*, Universidad Nacional de La Plata.
  <https://sedici.unlp.edu.ar/handle/10915/104778>

    The source for the five inherent characteristics being Accuracy,
    Completeness, Consistency, Credibility and Currentness — which is how the
    naming mismatch with *timeliness* was established — and for the
    decomposition of each characteristic into the fifteen quality properties
    and their identifiers implemented in this package.

* Demirhan, O. H. (2024). *Assessing Data Quality Management Issues in
  Open-Sourced Data Sets with Statistical Methodology.* MSc dissertation,
  University of Greenwich. Cites ISO/IEC 25024:2015 eight times, all within
  Chapter 2; the citation establishes what the dimensions mean and is not a
  claim of measure conformance.

The standard's Annexes A–E — the synoptic table of quality measure elements, the
measures indexed by element and target entity, and the alphabetic measure list —
are what would settle the remaining question of exactly which items belong in
each $A$ and $B$. They are not reproduced in any public source.
