---
title: "Methodology: every formula, stated"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodology: every formula, stated}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(datagrade)
```

Every score this package reports is written out below. Nothing is a black box,
and where a choice is arguable the argument is given rather than hidden.

All four dimension scores live on a **0--10 scale**. `as_pct()` converts once, at
the point of display, to 0--100. Holding one scale internally is deliberate: the
implementation this package replaces kept completeness on 0--10 and consistency
on 0--1 and multiplied both by ten, which is why its published consistency
figure for `iris` reads 2.5% where 25% was meant.

## Completeness

$$
\text{completeness} = 10 \times \left(1 - \frac{1}{p}\sum_{j=1}^{p} \frac{m_j}{n}\right)
$$

where $m_j$ is the number of missing values in column $j$, $n$ the row count and
$p$ the column count.

This is the **mean of the per-column missing rates**, not the overall share of
missing cells. The two differ whenever columns are unequally incomplete, and the
column-averaged form dilutes a badly damaged column: one column that is 30%
empty out of four costs 0.75 points, not 3. That is a property of the published
formula, kept here for continuity with the dissertation; if you want the cell
ratio, compute it from `report$missing$columns`.

```{r}
report <- dg_assess(airquality, verbose = FALSE)
report$missing$columns
report$scores[["completeness"]]
```

## Accuracy

$$
\text{accuracy} = 10 \times \left(1 - \frac{\text{outliers flagged}}{\text{numeric values tested}}\right)
$$

Outliers are flagged by one of two rules.

**Z-score.** A value is flagged when $\left|\frac{x_i - \mu}{\sigma}\right| > 3$.

**IQR.** A value is flagged when it falls outside
$[Q_1 - 1.5\,\text{IQR},\; Q_3 + 1.5\,\text{IQR}]$.

Two implementation choices are worth stating plainly.

*Outliers are counted on observed values, not imputed ones.* Mean imputation
adds mass at the centre of the distribution and shrinks $\sigma$, which
mechanically suppresses the very z-scores this step exists to find. The
dissertation ran this step after imputation; this package does not.

*A zero-variance column contributes zero outliers, not `NA`.* `scale()` returns
`NaN` when $\sigma = 0$, and summing that propagated `NA` through the entire
report, so a single constant column made the whole outlier count unreportable.

```{r}
df <- data.frame(constant = rep(5, 50), varying = c(rnorm(49), 100))
detect_outliers(df)$by_column
```

## Consistency

$$
\text{consistency} = 10 \times \frac{\left|\{j : \mathrm{VIF}_j \le \tau\}\right|}{p_{\text{num}}}
$$

with $\tau$ = `vif_threshold`, default 5.

The variance inflation factor is defined as $\mathrm{VIF}_j = \frac{1}{1 - R_j^2}$,
where $R_j^2$ comes from regressing column $j$ on all other numeric columns. This
package does not fit those regressions. It uses the identity that the VIFs are
the **diagonal of the inverted correlation matrix**:

$$
\mathrm{VIF} = \operatorname{diag}\left(\mathbf{R}^{-1}\right)
$$

which gives the same numbers in $O(p^3)$ time, independent of the row count. The
equivalence is asserted in the test suite against explicit `lm()` fits:

```{r}
set.seed(11)
d <- data.frame(a = rnorm(500), b = rnorm(500))
d$c <- 0.6 * d$a + 0.3 * d$b + rnorm(500)

dg_vif(d)

vapply(names(d), function(col) {
  r2 <- summary(lm(d[[col]] ~ ., data = d[setdiff(names(d), col)]))$r.squared
  1 / (1 - r2)
}, numeric(1))
```

Perfectly collinear columns are reported as `Inf` rather than dropped, so they
stay visible and still count against the score.

## Timeliness

$$
\text{timeliness} = 10 \times \left(1 - \frac{\text{future dates} + \text{stale dates}}{\text{dates present}}\right)
$$

A date is *future* if it is after `Sys.Date()`, *stale* if it is more than
`stale_years` (default 10) old.

This dimension is only measurable if date columns exist. Character columns are
therefore parsed as dates **before** any text normalisation runs -- otherwise
punctuation stripping turns `"2024-01-15"` into `"20240115"` and no date column
ever exists.

```{r}
df <- data.frame(when = c("2024-01-15", "2024-02-20"), value = 1:2)
r <- dg_assess(df, verbose = FALSE, strip_punctuation = TRUE)
r$parsed_dates
r$dates
```

## Distribution

Skewness and kurtosis use the sample moment forms:

$$
g_1 = \frac{1}{n}\sum \left(\frac{x_i-\mu}{\sigma}\right)^3
\qquad
g_2 = \frac{1}{n}\sum \left(\frac{x_i-\mu}{\sigma}\right)^4 - 3
$$

The $-3$ makes $g_2$ **excess** kurtosis, so a normal distribution scores 0. This
matters: the dissertation's formula subtracts 3 but its code called
`moments::kurtosis()`, which does not, so the published formula and the
published code disagreed by exactly 3. The formula is the contract here.

Normality uses the Shapiro-Wilk test, which is undefined above 5,000
observations. Larger columns are tested on a deterministic subsample and the
size actually used is reported in `shapiro_n`, so the result is never presented
as though the whole column were tested.

```{r}
set.seed(42)
analyze_distribution(data.frame(normal = rnorm(20000), skewed = rexp(20000)))
```

## The overall score

$$
\text{overall} = \frac{\sum_{d \in D} w_d \, s_d}{\sum_{d \in D} w_d}
$$

where $D$ is the set of dimensions that **apply** to the data. A data set with
no date columns cannot be scored on timeliness, so timeliness is dropped and the
remaining weights are renormalised rather than the data being penalised for a
dimension it cannot have.

```{r}
text_only <- data.frame(a = c("x", "y", "z"), stringsAsFactors = FALSE)
r <- dg_assess(text_only, verbose = FALSE)
r$scores
r$overall
```

## Relationship to ISO/IEC 25012 and 25024

The four dimension names are drawn from the ISO/IEC 25012 data quality model,
which is how the dissertation frames them. **The measures themselves are this
package's own and are not ISO/IEC 25024 conformant measures.** That claim is not
made anywhere in the package, and the reasons are worth being explicit about:

* ISO/IEC 25012's fourth inherent characteristic is **currentness**, not
  *timeliness*; *timeliness* comes from a different tradition. The package uses
  the dissertation's word.
* ISO/IEC 25024 measures take the form $X = A/B$ over counted *quality measure
  elements*. Computing $B$ generally requires metadata -- for the accuracy-range
  measure, $B$ is "the number of data items for which a required interval of
  values can be defined", which cannot be known from the data alone. This is
  precisely why the metadata-driven packages in this space require a data
  dictionary before they will run.
* ISO/IEC 25024 does not specify how measures aggregate into a per-characteristic
  score. The 0--10 weighted mean here is this package's choice, not a standard's.
* **Credibility**, the fifth inherent characteristic, is not measured at all.

Substituting a statistically derived interval for a specified one, as the
accuracy score does, measures *plausibility* rather than accuracy in the ISO
sense: an outlier need not be wrong, and a wrong value can sit exactly on the
mean. Read the accuracy score accordingly.

A future release will accept an optional specification object describing
expected ranges, required fields and valid domains. Supplying one would make the
genuine $X = A/B$ measures computable; omitting it would keep today's
zero-configuration behaviour.

### Sources for the claims in this section

ISO/IEC 25012 and 25024 are paywalled. Every statement above about their
*structure* is cited below; the individual measure identifiers and formulas are
deliberately **not** reproduced, because they could not be verified without the
documents themselves.

* ISO/IEC 25024:2015. *Systems and software engineering — Systems and software
  Quality Requirements and Evaluation (SQuaRE) — Measurement of data quality.*
  <https://www.iso.org/standard/35749.html>
* ISO/IEC 25012:2008. *Software engineering — SQuaRE — Data quality model.*
  <https://www.iso.org/standard/35736.html>
* Gualo, F., Rodriguez, M., Verdugo, J., Caballero, I. and Piattini, M. (2021).
  'Data Quality Certification using ISO/IEC 25012: Industrial Experiences',
  *Journal of Systems and Software*. <https://arxiv.org/abs/2102.11527>

    The source for three specific claims: that ISO/IEC 25024 measures take the
    form $X = A/B$ on a ratio scale over $[0.0, 1.0]$; that the Accuracy Range
    measure (RAN_EXAC) defines its element $B$ as "the number of data items for
    which can be defined a required interval of values"; and that "ISO/IEC 25024
    does not specifically address how the measures corresponding to the data
    quality properties should be aggregated to compute the overall quality level
    for each data quality characteristic." The authors are from AQCLab, an
    accredited ISO/IEC 25012 data quality evaluation laboratory.

* Calabrese, J., Esponda, S. and Pesado, P. (2020). 'Framework for Data Quality
  Evaluation Based on ISO/IEC 25012 and ISO/IEC 25024', *8th Conference on Cloud
  Computing, Big Data & Emerging Topics*, Universidad Nacional de La Plata.
  <https://sedici.unlp.edu.ar/handle/10915/104778>

    The source for the five inherent characteristics being Accuracy,
    Completeness, Consistency, Credibility and Currentness — which is how the
    naming mismatch with *timeliness* was established.

* Demirhan, O. H. (2024). *Assessing Data Quality Management Issues in
  Open-Sourced Data Sets with Statistical Methodology.* MSc dissertation,
  University of Greenwich. Cites ISO/IEC 25024:2015 eight times, all within
  Chapter 2; the citation establishes what the dimensions mean and is not a
  claim of measure conformance.

The standard's Annexes A–E — the synoptic table of quality measure elements, the
measures indexed by element and target entity, and the alphabetic measure list —
are what a conformant implementation would need. They are not reproduced in any
public source.
